Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-09 chapters/chapter02.tex
fixing floats
comments
some revsion edits
implementing comments

%---CONVOLUTION ILLUSTRATION---%
\begin{figure}[H]
\centering
\begingroup
\everymath{\displaystyle}
{\Large
\begin{displaymath} % 
    \xymatrix{ 
    \bullet^{t_u} \ar@{.>}[rr]_{\mathbf{P}_{1}(t_{u},T_{1},r_v)} && \bullet^{T_1} \ar@{.>}[rr]_{\mathbf{P}_{2}(T_{1},t_{v},r_v)} && \bullet^{t_v} \\
    \bullet^{t_u} \ar@{.>}[rrrr]_{\mathbf{P}(t_{u},t_{v},r_v)} & & & & \bullet^{t_v}
    } % 
\end{displaymath}
}
\endgroup
\caption{
{ \footnotesize 
{\bf Collapsing branches.} Transition probability matrix $\mathbf{P}(t_u,t_v,r_v)$ governs the inhomogeneous substitution process along a branch from time $t_u$ to $t_v$ and is the matrix-product of transition matrices $\mathbf{P}_{1}(t_u,T_1,r_v)$ and $\mathbf{P}_{2}(T_1,t_v,r_v)$, where $T_1$ is the epoch change-point time between homogeneous processes 1 and 2.  We assume rate scalar $r_v$ remains constant along the entire branch.
} % END: footnotesize
}
\label{fig:collapsing}
\end{figure}

@@

%---UPDATE-CONVOLVE INFOGRAPHIC---%
\begin{figure}[H]
\centering
\includegraphics[scale=0.35]{infographic} 
\caption{
{ \footnotesize 
{\bf A small-scale example of the update convolve routine for a 5-taxon tree.} 
In this example, transition times are set at $t_{1} = 1.0$ and $t_{2} = 4.5$, creating three respective epochs indicated by alternating background.
Black dots represent transition probability buffers for particular branches, grey dots denote extra buffers allocated for partial transition matrices.
We allocate 10 extra buffers.  
Consecutive panels illustrate the order of update and convolve steps in the routine.
Update step in A adds buffers to the update work queue. 
In B and C, the routine starts adding buffers to the convolve queue in a stepwise fashion, dictated by the tree traversal, freeing extra buffers whenever the queue is dispatched for execution. 
In D, the routine uses freed buffers for a new update step.
In E, the parallel convolve queue is used to compute the final transition matrix stored in buffer 5, concluding the work.
}% END: footnotesize
}
\label{fig:UpdateConvolveExample}
\end{figure}

@@

%---TOPOLOGY FOR SIMULATIONS---%
\begin{figure}[H]
\centering
% \color{fgcolor}
\includegraphics[scale=0.35]{treeplot} 
\caption{
{ \footnotesize 
{\bf  Epoch simulation scenarios on an influenza A maximum clade credibility tree topology.} 
% \myedit{DPeight}{
In the two-epoch example illustrated by the top alternating background
% at the top, 
the transition time is set at $t_{1}=7$, creating two epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$ and $\mathbf{Q_{2}}$ respectively, separated by the light and dark grey areas and the dotted line.
In the three-epoch example illustrated 
by the bottom alternating background
% at the bottom, 
transition times are put at $t_{1}=7$ and $t_{2}=15$, creating three epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$, $\mathbf{Q_{2}}$ and then again $\mathbf{Q_{1}}$, as indicated by the alternating dark and light areas and the dotted lines.
% }%END: myedit
}% END: footnotesize
}%END: caption
\label{fig:2RateTree}
\end{figure}

In Figure~\ref{fig:2RateTree}, we illustrate the tree topology and the transition times defined for both a two-epoch and three-epoch specification.
This tree has 69 tips, is rooted and time-scaled, effectively covering a period of about 18 years. 
For each replicate data set, we simulate $1000$ nucleotide sites under the Hasegawa, Kishino and Yano (HKY) model \citep{hky85}. 
We set the substitution rates and base frequencies to values estimated from the real data set.

@@

%---SHANKARAPPA ANALYSIS RESULTS HPD---%
\begin{figure}[H]
\centering
% \color{fgcolor}
\includegraphics[scale=0.6]{shankarappa_hpd_new} 
\caption{
{ \footnotesize 
{\bf Estimates of $d_N/d_S$ ratio for within-host HIV analyses.} Vertical lines represent 95\% highest posterior density intervals for the $d_N/d_S$ ratio estimates. 
Parameter $\omega$ is estimated under the homogeneous model, while $\omega_{1}$ and $\omega_{2}$ are obtained using the epoch model.
}% END: footnotesize
}
\label{fig:shank_hpd}
\end{figure}

@@

%---SHANKARAPPA ANALYSIS RESULTS BF---%
\begin{table}[H]
\centering
% \medskip
% \begin{minipage}{\textwidth} 
% \footnotesize{
\begin{tabular}%{ccc}
{@{}l*{2}{D{.}{.}{7}}@{}}
\hline 
\textbf{Patient} & \textbf{Posterior probability} & \textbf{log Bayes factor}\tabularnewline
\hline 
patient 1 & $0.999$ & $7.418$\tabularnewline
patient 2 & >$0.999$ & $9.602$\tabularnewline
patient 3 & $0.898$ & $2.174$\tabularnewline
patient 5 & $0.430$ & $-0.282$\tabularnewline
patient 6 & >$0.999$ & $9.210$\tabularnewline
patient 7 & >$0.999$ & $8.112$\tabularnewline
patient 8 & $0.933$ & $2.627$\tabularnewline
patient 9 & $0.895$ & $2.142$\tabularnewline
\hline 
\textbf{Joint evidence:} & $0.894$ & $2.14$ \tabularnewline
\end{tabular}
% } % END: footnotesize
\caption{
{ \footnotesize 
{\bf Bayes factor test for decreased selection after progression.} We report the posterior probability that $\omega_{1} < \omega_{2}$ and the corresponding Bayes factor against the alternative that $\omega_{1} \ge \omega_{2}$.
}% END: footnotesize
}
\label{tab:shank_bf}
% \end{minipage}
\end{table}

@@

In a second application of the epoch model, we focus on discrete diffusion processes to infer spatio-temporal history from viral gene sequences.
This type of phylogeographic inference, where the sampling locations are considered as discrete geographic traits, has gained popularity in recent years, at least partly because of a flexible and efficient Bayesian implementation that connects dispersal dynamics to sequence evolution in time-measured phylogenies \citep{Lemey2009}.
Recently \citet{Bahl2011} have applied this Bayesian inference framework to investigate the circulation dynamics of global influenza A H3N2 through time.
Since the authors were interested in capturing the heterogeneity in these dynamics over successive seasonal epidemics between 2003 and 2006, they consider discrete traits that are the product of sampling location and sampling time (epidemic season).
Not only does this discretization by sampling time seem counterintuitive for a model that emits discrete outcomes as a continuous function of time, it also considerably increases the dimensionality of the CTMC rate matrix and thus the number of parameters to inform by the sparse spatial data.

@@

%---FLU DATA RESULTS MAP---%
\begin{figure}[H]
\centering
\includegraphics[scale=0.5]{flu2_epoch} 
\caption{
{ \footnotesize 
{\bf A two-epoch phylogeographic model applied to seasonal influenza H3N2.} 
A. Maximum clade credibility (MCC) tree with branches colored according to modal discrete location states at each node. The grey time intervals represent the epoch model with a single discrete rate matrix shared across northern hemisphere spring and summer (light grey) time intervals and another rate matrix shared across the northern hemisphere autumn and winter (dark grey) time intervals. B. Diffusion rates supported by a Bayes factor $>20$ for spring and summer epoch intervals. The width of the arrows reflects the magnitude of the Bayes factor support. C. Diffusion rates supported by a Bayes factor $>20$ for autumn and winter epoch intervals.
}% END: footnotesize 
}
\label{fig:2EpochFlu}
\end{figure}

@@

%---FLU DATA RESULTS PS SS---%
% \afterpage{\clearpage
\begin{table}[H]
\begin{minipage}{\textwidth} 
\centering
% \footnotesize{
\begin{tabular}{ccc}
\hline 
\multirow{2}{*}{\textbf{Model}} & \multicolumn{2}{c}{\textbf{Marginal likelihood}}\tabularnewline
 & PS\footnote{Path Sampling} & SS\footnote{Stepping Stone Sampling}\tabularnewline
\hline 
homogeneous & -827.29 & -825.07\tabularnewline
7-epoch & -806.36  & -803.40 \tabularnewline
14-epoch & -798.77  & -795.63 \tabularnewline
\end{tabular}
% } % END: footnotesize
\caption{
{ \footnotesize 
{\bf Marginal likelihood estimates.} Comparison in terms of model fit between a homogeneous model, an epoch model with time discretized into $S=7$ epochs alternating between 2 different rate matrices and an epoch model with time discretized into $S=14$ epochs, alternating between 4 separate rate matrices.
} % END: footnotesize
}
\label{tab:flu_ps}
\end{minipage}
\end{table}
% \clearpage }% END: afterpage

@@

%---STACK SIZE TIMING---%
\begin{figure}[H]
\centering
\includegraphics[scale=0.6]{timing} 
\caption{
{\footnotesize 
{\bf Relative speed-ups as a function the number of extra buffers allocated for two real data examples.}
Speed-ups are reported relative to the serial execution, using system time spent sampling 1,000,000 states from the posterior distribution using an MCMC sampler. 
Black dots denote update transition matrices operations, grey dots denote convolve operations.
A. Speed-ups for a flu data set (522 taxa, discrete geographic diffusion process between $K=7$ states, time discretized into $S=13$ epochs, see {\it{Seasonal circulation dynamics of human influenza A}}). 
B. Speed-ups for a patient 3 from HIV data set (109 taxa, 516 sites, state-space size of $K=61$ codons, $S=2$ epochs, see {\it{Within-host HIV selection dynamics}}). 
} % END: footnotesize 
}
\label{fig:StackSizeTiming}
\end{figure}

>>>>

%---CONVOLUTION ILLUSTRATION---%
\begin{figure}[h!]
\centering
\begingroup
\everymath{\displaystyle}
{\Large
\begin{displaymath} % 
    \xymatrix{ 
    \bullet^{t_u} \ar@{.>}[rr]_{\mathbf{P}_{1}(t_{u},T_{1},r_v)} && \bullet^{T_1} \ar@{.>}[rr]_{\mathbf{P}_{2}(T_{1},t_{v},r_v)} && \bullet^{t_v} \\
    \bullet^{t_u} \ar@{.>}[rrrr]_{\mathbf{P}(t_{u},t_{v},r_v)} & & & & \bullet^{t_v}
    } % 
\end{displaymath}
}
\endgroup
\caption{
{ \footnotesize 
{\bf Collapsing branches.} Transition probability matrix $\mathbf{P}(t_u,t_v,r_v)$ governs the inhomogeneous substitution process along a branch from time $t_u$ to $t_v$ and is the matrix-product of transition matrices $\mathbf{P}_{1}(t_u,T_1,r_v)$ and $\mathbf{P}_{2}(T_1,t_v,r_v)$, where $T_1$ is the epoch change-point time between homogeneous processes 1 and 2.  We assume rate scalar $r_v$ remains constant along the entire branch.
} % END: footnotesize
}
\label{fig:collapsing}
\end{figure}

@@

%---UPDATE-CONVOLVE INFOGRAPHIC---%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.35]{infographic} 
\caption{
{ \footnotesize 
{\bf A small-scale example of the update convolve routine for a 5-taxon tree.} 
In this example, transition times are set at $t_{1} = 1.0$ and $t_{2} = 4.5$, creating three respective epochs indicated by alternating background.
Black dots represent transition probability buffers for particular branches, grey dots denote extra buffers allocated for partial transition matrices.
We allocate 10 extra buffers.  
Consecutive panels illustrate the order of update and convolve steps in the routine.
Update step in A adds buffers to the update work queue. 
In B and C, the routine starts adding buffers to the convolve queue in a stepwise fashion, dictated by the tree traversal, freeing extra buffers whenever the queue is dispatched for execution. 
In D, the routine uses freed buffers for a new update step.
In E, the parallel convolve queue is used to compute the final transition matrix stored in buffer 5, concluding the work.
}% END: footnotesize
}
\label{fig:UpdateConvolveExample}
\end{figure}

@@

%---TOPOLOGY FOR SIMULATIONS---%
\begin{figure}[h!]
\centering
% \color{fgcolor}
\includegraphics[scale=0.35]{treeplot} 
\caption{
{ \footnotesize 
{\bf  Epoch simulation scenarios on an influenza A maximum clade credibility tree topology.} 
% \myedit{DPeight}{
In the two-epoch example illustrated by the top alternating background
% at the top, 
the transition time is set at $t_{1}=7$, creating two epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$ and $\mathbf{Q_{2}}$ respectively, separated by the light and dark grey areas and the dotted line.
In the three-epoch example illustrated 
by the bottom alternating background
% at the bottom, 
transition times are put at $t_{1}=7$ and $t_{2}=15$, creating three epochs with substitution processes governed by infinitesimal rate matrices $\mathbf{Q_{1}}$, $\mathbf{Q_{2}}$ and then again $\mathbf{Q_{1}}$, as indicated by the alternating dark and light areas and the dotted lines.
% }%END: myedit
}% END: footnotesize
}%END: caption
\label{fig:2RateTree}
\end{figure}

In Figure~\ref{fig:2RateTree}, we illustrate the tree topology and the transition times defined for both a two-epoch and three-epoch specification.
This tree has 69 tips, is rooted and time-scaled, effectively covering a period of about 18 years. 
For each replicate data set, we simulate $1000$ nucleotide sites under the Hasegawa, Kishino and Yano (HKY) model \citep{hky85}. 
We set the substitution rates and base frequencies to values estimated from the real data set.
\myedit{svtwo}{
In all our analyses of the replicate data, as well as in the analyses of the real-world data sets in the next sections, we consider the tree as random and estimate it from the nucleotide data.
}

@@

%---SHANKARAPPA ANALYSIS RESULTS HPD---%
\begin{figure}[h!]
\centering
% \color{fgcolor}
\includegraphics[scale=0.6]{shankarappa_hpd_new} 
\caption{
{ \footnotesize 
{\bf Estimates of $d_N/d_S$ ratio for within-host HIV analyses.} Vertical lines represent 95\% highest posterior density intervals for the $d_N/d_S$ ratio estimates. 
Parameter $\omega$ is estimated under the homogeneous model, while $\omega_{1}$ and $\omega_{2}$ are obtained using the epoch model.
}% END: footnotesize
}
\label{fig:shank_hpd}
\end{figure}

@@

%---SHANKARAPPA ANALYSIS RESULTS BF---%
\begin{table}[h!]
\centering
% \medskip
% \begin{minipage}{\textwidth} 
% \footnotesize{
\begin{tabular}%{ccc}
{@{}l*{2}{D{.}{.}{7}}@{}}
\hline 
\textbf{Patient} & \textbf{Posterior probability} & \textbf{log Bayes factor}\tabularnewline
\hline 
patient 1 & $0.999$ & $7.418$\tabularnewline
patient 2 & >$0.999$ & $9.602$\tabularnewline
patient 3 & $0.898$ & $2.174$\tabularnewline
patient 5 & $0.430$ & $-0.282$\tabularnewline
patient 6 & >$0.999$ & $9.210$\tabularnewline
patient 7 & >$0.999$ & $8.112$\tabularnewline
patient 8 & $0.933$ & $2.627$\tabularnewline
patient 9 & $0.895$ & $2.142$\tabularnewline
\hline 
\textbf{Joint evidence:} & $0.894$ & $2.14$ \tabularnewline
\end{tabular}
% } % END: footnotesize
\caption{
{ \footnotesize 
{\bf Bayes factor test for decreased selection after progression.} We report the posterior probability that $\omega_{1} < \omega_{2}$ and the corresponding Bayes factor against the alternative that $\omega_{1} \ge \omega_{2}$.
}% END: footnotesize
}
\label{tab:shank_bf}
% \end{minipage}
\end{table}

@@

In a second application of the epoch model, we focus on discrete diffusion processes to infer spatio-temporal history from viral gene sequences.
This type of phylogeographic inference, where the sampling locations are considered as discrete geographic traits, has gained popularity in recent years, at least partly because of a flexible and efficient Bayesian implementation that connects dispersal dynamics to sequence evolution in time-measured phylogenies \citep{Lemey2009}.
Recently \citet{Bahl2011} have applied this Bayesian inference framework to investigate the circulation dynamics of global influenza A H3N2 through time.
Since the authors were interested in capturing the heterogeneity in these dynamics over successive seasonal epidemics between 2003 and 2006, they consider discrete traits that are the product of sampling location and sampling time (epidemic season).
\myedit{svsixteen}{
This means they assumed that each geographic region and influenza season is considered to be a discrete character, for example samples collected from October 2002 to February 2003 in New York as assigned discrete state labelled ``New York 2003'' \citep{Bahl2011}.
}
Not only does this discretization by sampling time seem counterintuitive for a model that emits discrete outcomes as a continuous function of time, it also considerably increases the dimensionality of the CTMC rate matrix and thus the number of parameters to inform by the sparse spatial data.

@@

%---FLU DATA RESULTS MAP---%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.5]{flu2_epoch} 
\caption{
{ \footnotesize 
{\bf A two-epoch phylogeographic model applied to seasonal influenza H3N2.} 
A. Maximum clade credibility (MCC) tree with branches colored according to modal discrete location states at each node. The grey time intervals represent the epoch model with a single discrete rate matrix shared across northern hemisphere spring and summer (light grey) time intervals and another rate matrix shared across the northern hemisphere autumn and winter (dark grey) time intervals. B. Diffusion rates supported by a Bayes factor $>20$ for spring and summer epoch intervals. The width of the arrows reflects the magnitude of the Bayes factor support. C. Diffusion rates supported by a Bayes factor $>20$ for autumn and winter epoch intervals.
}% END: footnotesize 
}
\label{fig:2EpochFlu}
\end{figure}

@@

%---FLU DATA RESULTS PS SS---%
% \afterpage{\clearpage
\begin{table}[h!]
\begin{minipage}{\textwidth} 
\centering
% \footnotesize{
\begin{tabular}{ccc}
\hline 
\multirow{2}{*}{\textbf{Model}} & \multicolumn{2}{c}{\textbf{Marginal likelihood}}\tabularnewline
 & PS\footnote{Path Sampling} & SS\footnote{Stepping Stone Sampling}\tabularnewline
\hline 
homogeneous & -827.29 & -825.07\tabularnewline
7-epoch & -806.36  & -803.40 \tabularnewline
14-epoch & -798.77  & -795.63 \tabularnewline
\end{tabular}
% } % END: footnotesize
\caption{
{ \footnotesize 
{\bf Marginal likelihood estimates.} Comparison in terms of model fit between a homogeneous model, an epoch model with time discretized into $S=7$ epochs alternating between 2 different rate matrices and an epoch model with time discretized into $S=14$ epochs, alternating between 4 separate rate matrices.
} % END: footnotesize
}
\label{tab:flu_ps}
\end{minipage}
\end{table}
% \clearpage }% END: afterpage

@@

%---STACK SIZE TIMING---%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.6]{timing} 
\caption{
{\footnotesize 
{\bf Relative speed-ups as a function the number of extra buffers allocated for two real data examples.}
Speed-ups are reported relative to the serial execution, using system time spent sampling 1,000,000 states from the posterior distribution using an MCMC sampler. 
Black dots denote update transition matrices operations, grey dots denote convolve operations.
A. Speed-ups for a flu data set (522 taxa, discrete geographic diffusion process between $K=7$ states, time discretized into $S=13$ epochs, see {\it{Seasonal circulation dynamics of human influenza A}}). 
B. Speed-ups for a patient 3 from HIV data set (109 taxa, 516 sites, state-space size of $K=61$ codons, $S=2$ epochs, see {\it{Within-host HIV selection dynamics}}). 
} % END: footnotesize 
}
\label{fig:StackSizeTiming}
\end{figure}
